---
title: "Lasso Prediction Model"
output: pdf_document
date: "2024-05-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
insurance = read.csv("insurance.csv")
summary(insurance)
```


```{r}
sum(is.na(insurance))
insurance_clean = na.omit(insurance)
insurance_filled = insurance
insurance_filled$age = ifelse(is.na(insurance_filled$age), mean(insurance_filled$age, na.rm = TRUE), insurance_filled$age)
```


```{r}
library(glmnet)
library(dplyr)
insurance_data <- insurance %>%
  mutate(across(c(sex, smoker, region), as.factor)) %>%
  mutate(across(c(sex, smoker, region), as.numeric))
x <- as.matrix(insurance_data[, -7])
y <- insurance_data$charges
```
The second model trained will be a Lasso regression model. We will start by encoding the categorical varibales in order to proceed and creating x and y.


```{r}
set.seed(100) 
model <- glmnet(x, y, alpha = 1)  
cross_validation <- cv.glmnet(x, y, alpha = 1)
lambda <- cross_validation$lambda.min
lasso_model <- glmnet(x, y, alpha = 1, lambda = lambda)
coef(lasso_model)
```
We'll be using cross validation in order to choose the optimal value for lambda and setting a seed to ensure reproducibility of the code. 


```{r}
set.seed(100)
num_rows <- nrow(insurance_data)
training_rows <- floor(0.8 * num_rows)
training_indices <- sample(seq_len(num_rows), size = training_rows)
training_data <- insurance_data[train_indices,]
testing_data <- insurance_data[-train_indices,]
x_train <- as.matrix(training_data[, -7])  
x_test <- as.matrix(testing_data[, -7])
y_train <- training_data$charges
y_test <- testing_data$charges
```
We'll be creating training and testing datasets by randomly selecting 80% of the original dataset to be the training dataset and the remaining 20% to be the testing dataset.


```{r}
training_model <- glmnet(x_train, y_train, alpha = 1, lambda = lambda)
predictions <- predict(training_model, s = lambda, newx = x_test)
rmse <- sqrt(mean((predictions - y_test)^2))
rsq <- cor(y_test, predictions)^2

plot(y_test, predictions, main = "Charges and Predicted Charges",
     xlab = "Charges", ylab = "Predicted Charges",
     pch = 19, col = 'brown')
abline(0, 1, col = "blue")
rsq <- cor(y_test, predictions)^2
rsq
```

We can see that there is a strong correlation between the predictors and the medical charges one will incurr.